A data-driven model of nucleosynthesis with chemical tagging in latent space

### 2019-05-20 article/mcfa.tex
minor

\begin{abstract}
Chemical tagging seeks to identify unique star formation sites from
present-day stellar abundances.
Previous techniques have treated each abundance dimension as being
statistically independent, despite theoretical expectations that
many elements can be produced by more than one nucleosynthetic process.
In this work we introduce a data-driven model of nucleosynthesis 
where a set of latent factors (e.g., nucleosynthetic yields) contribute
to all stars with different scores, and clustering (e.g., chemical tagging) 
is modelled by a mixture of multivariate gaussians
in lower-dimensional latent space.
We use an exact method to simultaneously estimate the factor scores for
each star, the partial assignment of each star
to each cluster, and the latent factors common to all stars.
We use an information-theoretic Bayesian principle to estimate the number of
latent factors and clusters.
We find that five latent factors are preferred to explain 18 chemical elements
in the second \Galah\ data release.
From these latent factors we identify the rapid- and slow-neutron capture 
processes, as well as latent factors consistent with Fe-peak and 
$\alpha$-element production, and a fifth latent factor where only K and Zn
dominate. We discuss examples where the inferred latent factors deviate from 
expected yields, which includes an  indication that Sc is associated with the r-process.
We find three clusters are preferred given 1,072 stars in \Galah,
two of which are representative of the [$\alpha$/Fe]-poor and 
[$\alpha$/Fe]-rich populations in the Milky Way disk. 
We release accompanying software that scales well with the available data,
allowing for model parameters to be optimised in seconds given a fixed number
of latent factors, components, and  $\sim3 \times 10^7$ abundance measurements.
\end{abstract}

@@

In this experiment we perform blind chemical tagging using the 
photospheric abundances released as part of the second \Galah\ 
data release \citep{Buder:2018}. This data set includes
up to 23 chemical abundances reported for 342,682 stars.
Some abundances are not reported for a variety of 
reasons related to astrophysics or the adopted analyses
(e.g., a star is too hot to estimate the abundance for a given
electronic transition, or the signal-to-noise of the spectrum is too low).
Our method as described cannot efficiently handle missing data 
entries for a given star, and as such we are forced to exclude stars that
do not have a full complement of abundances. However, we also
chose to exclude some elemental abundances from our chemical
tagging experiments. For example, our experiments
with \Galah\ data do not include lithium or carbon abundances because the
photospheric varies throughout a star's lifetime (e.g., see Experiment 2). This is true
to a small degree for many elements \citep[e.g.,][]{Dotter:2017},
but for the purposes of this experiment we assume that all other
photospheric abundances remain constant throughout a star's
lifetime.

@@

We first selected stars with \texttt{flag\_cannon = 0} to exclude
stars where there is reason to suspect that the stellar parameters
(e.g., $\teff$, $\logg$) are unreliable, and as a result the 
detailed chemical abundances would be untrustworthy. We then took all
stars with a signal-to-noise ratio exceeding 40 in the blue arm (\texttt{snr\_c1} $>$ 40), and
stars with no erroneous flags in all of the following abundances: Na, Al, Si, K, Ca, Sc, Ti, V, Mn, Fe, Ni, Cu, Zn, Y, Ba, La, and Eu.
%These elements are produced from multiple nucleosynthetic
%pathways, and 
%It is this subsample that we will use for 
%our initial chemical experiment with \Galah\ data.
These elements were chosen because they trace multiple nucleosynthetic pathways, and they are more commonly reported in the \Galah\ data release, allowing for a larger number of stars with a complete abundance inventory.
There are 1,072 stars that met these criteria.

@@

% This paper was built from reference XXX on github repo YYY

@@

\software{
	\package{Astropy}\,\,\citep{astropy:v1,astropy:v2},\,\,
    \package{IPython}\,\,\citep{ipython},\,\,
    \package{matplotlib}\,\,\citep{mpl},\,\,
    \package{numpy}\,\,\citep{numpy},\,\,
    \package{scipy}\,\,\citep{scipy},\,\,
%    \package{Stan}\,\,\citep{stan},\,\,
    \package{Jupyter Notebooks}\,\,\citep{jupyter-notebooks}\,\,
}    % TOPCAT

>>>>

\begin{abstract}
Chemical tagging seeks to identify unique star formation sites from
present-day stellar abundances.
Previous techniques have treated each abundance dimension as being
statistically independent, despite theoretical expectations that
many elements can be produced by more than one nucleosynthetic process.
In this work we introduce a data-driven model of nucleosynthesis 
where a set of latent factors (e.g., nucleosynthetic yields) contribute
to all stars with different scores, and clustering (e.g., chemical tagging) 
is modelled by a mixture of multivariate gaussians
in lower-dimensional latent space.
We use an exact method to simultaneously estimate the factor scores for
each star, the partial assignment of each star
to each cluster, and the latent factors common to all stars.
We use an information-theoretic Bayesian principle to estimate the number of
latent factors and clusters.
We find that five latent factors are preferred to explain 18 chemical elements
in the second \Galah\ data release.
From these latent factors we identify the rapid- and slow-neutron capture 
processes, as well as latent factors consistent with Fe-peak and 
$\alpha$-element production, and a fifth latent factor where only K and Zn
dominate. We discuss examples where the inferred latent factors deviate from 
expected yields, which includes an  indication that Sc is associated with the r-process.
We find three clusters are preferred given 1,072 stars in \Galah,
two of which are representative of the [$\alpha$/Fe]-poor and 
[$\alpha$/Fe]-rich populations in the Milky Way disk. 
We release accompanying software that scales well with the available data,
allowing for model parameters to be optimised in seconds given a fixed number
of latent factors, components, and $\sim10^6$ abundance measurements.
\end{abstract}

@@

In this experiment we perform blind chemical tagging using the 
photospheric abundances released as part of the second \Galah\ 
data release \citep{Buder:2018}. This data set includes
up to 23 chemical abundances reported for 342,682 stars.
Some abundances are not reported for a variety of 
reasons related to astrophysics or the adopted analyses
(e.g., a star is too hot to estimate the abundance for a given
electronic transition, or the signal-to-noise of the spectrum is too low).
Our method as described cannot efficiently handle missing data 
entries for a given star, and as such we are forced to exclude stars that
do not have a full complement of abundances. However, we also
chose to exclude some elemental abundances from our chemical
tagging experiments. For example, our experiments
with \Galah\ data do not include lithium or carbon abundances because the
photospheric values will vary throughout a star's lifetime \citep[e.g.,][]{Casey:2016b,Casey:2019}. This is true
to a small degree for many elements \citep[e.g.,][]{Dotter:2017},
but for the purposes of this experiment we assume that all other
photospheric abundances remain constant throughout a star's
lifetime.

@@

We first selected stars with \texttt{flag\_cannon = 0} to exclude
stars where there is reason to suspect that the stellar parameters
(e.g., $\teff$, $\logg$) are unreliable, and as a result the 
detailed chemical abundances would be untrustworthy. We then took all
stars with a signal-to-noise ratio exceeding 40 in the blue arm (\texttt{snr\_c1 > 40}), and
stars with no erroneous flags in all of the following abundances: Na, Al, Si, K, Ca, Sc, Ti, V, Mn, Fe, Ni, Cu, Zn, Y, Ba, La, and Eu.
%These elements are produced from multiple nucleosynthetic
%pathways, and 
%It is this subsample that we will use for 
%our initial chemical experiment with \Galah\ data.
These elements were chosen because they trace multiple nucleosynthetic pathways, and they are more commonly reported in the \Galah\ data release, allowing for a larger number of stars with a complete abundance inventory.
There are 1,072 stars that met these criteria.

@@

% TODO:
%\vspace{2em}
%This \emph{Article} was compiled from the \texttt{git} commit \todo{\texttt{HASH}}.
%The analysis, results, and figures presented in this \emph{Article} can be reproduced in full 
%using the following commands in a modern terminal:\vspace{1em}\\
%\texttt{git clone https://github.com/andycasey/mcfa}\\
%\texttt{cd mcfa/}\\
%\todo{\texttt{sh article/build.sh}}

@@

\clearpage

\software{
	\package{Astropy} \citep{astropy:v1,astropy:v2},\,\,
    \package{IPython} \citep{ipython},\,\,
    \package{matplotlib} \citep{mpl},\,\,
    \package{numpy} \citep{numpy},\,\,
    \package{scipy} \citep{scipy},\,\,
%    \package{Stan}\,\,\citep{stan},\,\,
    \package{Jupyter Notebooks} \citep{jupyter-notebooks}
}    % TOPCAT
